Optimizing Abstract Abstract Machines

### 2012-11-13 paper/paper.tex
Incorporating Matt's comments.

\begin{align*}
\mathit{push}_\mlab^\mcntr(\msto,\mkont) &=
  \mlab\mcntr,\msto\sqcup[\mlab\mcntr\mapsto\{\mkont\}] \\
\mathit{bind}^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto \maddr],
                                           \msto\sqcup[\maddr \mapsto
                                             \{ \mval\}],
                                           \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= x\mcntr' \\
%%              \lfloor \mtcntr \rfloor_k &= \mtcntr \\
              \lfloor \mcntr \rfloor_0 &= \mtcntr \\
              \lfloor \mlab\mcntr \rfloor_{k+1} &= \mlab\lfloor \mcntr\rfloor_k
\end{align*}
For the duration of this paper, since our example abstract value space
has no interesting ordering, $\sqcup$ is a point-wise lifting of
$\cup$:
\begin{align*}
\msto\sqcup \varnothing &= \msto \\
\msto\sqcup \msto'[\maddr \mapsto S] &= \lambda \maddralt. \left\{
  \begin{array}{ll}
    \msto(\maddr)\cup S & \mbox{if } \maddralt = \maddr, \maddr\in dom(\msto) \\
    S & \mbox{if } \maddralt = \maddr, \maddr\notin dom(\msto) \\
    (\msto\sqcup\msto')(\maddralt) & \mbox{otherwise}
  \end{array}\right.
\end{align*}

\paragraph{Primitives}

@@

The uniform $k$-CFA allocation strategy would make the fixpoint of the
reduction relation in figure \ref{fig:aam} a computable analysis, but
not an effective one. This is not the strategy that AAM, nor we,
recommend. We explain a succession of approximations to reach the
baseline analysis that we compare performance against. But first, we
need a fixpoint combinator to drive the succession of reduction
relations we define.

@@

We compute the semantics of a program by iterating the state
transition relation until a fixed point in the reachable states is
reached. We begin with a generic fixpoint combinator, ${\mathcal F}$:

@@

A common technique to faster convergence in flow analyses is to share
a common store that is monotonically updated as interpretation
continues. In the ISWIM language with only unary operators, this makes
our 0-CFA quartic in the size of the program. Not great, but
polynomial.

@@

Let $wn : \widehat{State} \times Store \to State$ be the metafunction
that adds a store to a storeless state and $c.\msto$ projects the
store out of a state $c$.

\begin{align*}
(cs, \msto) &\widehat{\machstep} (cs', \msto') \\
\mbox{where } cs' &= \{c' \mid wn(c, \msto) \machstep wn(c, \msto^c), c \in cs\} \\
              \msto' &= \bigsqcup\{\msto^c \mid wn(c,\msto)\machstep wn(c', \msto^c), c \in cs\}
\end{align*}
That is, the next set of states is what we get by stepping each
given state, without its store. The next store is the combination of all produced
stores. For simplicity, assume that stuck states step to themselves.

\subsection{Baseline evaluation}
\label{sec:baselineeval}

@@

To achieve this, we instead allocate an address for the value position
when we create the continuation.  We will refer to this change in the
semantics as store-allocated results. This address and the tail address are
both determined by the label of the application point, so the space
becomes linear and the overall complexity drops to cubic. Note that
this is a critical abstraction in languages with n-ary functions,
since otherwise the continuation space becomes ${\mathcal O}(n^n)$!

@@

This extra store-allocation is effectively naming all intermediate
results, and thus the precision aligns with an analysis specialized to
ANF. We can also play a representation trick and remove
$\menv$ since we know it is the identity function in 0-CFA.

@@

\begin{align*}
(cs, \msto) &\widehat{\machstep} (cs\cup cs', \msdiff(\msto)) \\
\mbox{ where } (cs', \msdiff) &= step^*(\varnothing, cs, \mtlst) \\
step^*(S, \varnothing, \msdiff) &= (S, \msdiff) \\
step^*(S, \{c\}\cup cs, \msdiff &= step^*(S\cup cs^*, cs, \msdiff^*) \\
cs^* &= \{c' \mid (c,\msto,\msdiff) \machstep (c',\msdiff^c) \} \\
\msdiff^* &= concat\{\msdiff^c \mid (c, \msto,\msdiff) \machstep (c', \msdiff^c)\}
\end{align*}

>>>>

\begin{align*}
\mathit{push}_\mlab^\mcntr(\msto,\mkont) &=
  \mlab\mcntr,\msto\sqcup[\mlab\mcntr\mapsto\{\mkont\}] \\
\mathit{bind}^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto \maddr],
                                           \msto\sqcup[\maddr \mapsto
                                             \{ \mval\}],
                                           \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= x\mcntr' \\
%%              \lfloor \mtcntr \rfloor_k &= \mtcntr \\
              \lfloor \mcntr \rfloor_0 &= \mtcntr \\
              \lfloor \mlab\mcntr \rfloor_{k+1} &= \mlab\lfloor \mcntr\rfloor_k
\end{align*}
$\sqcup$ on stores is a point-wise lifting of $\sqcup$: $\msto \sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup \msto'(\maddr)$.
For the duration of this paper, since our example abstract value space
has no interesting ordering, $\sqcup$ can be read as $\cup$.
\paragraph{Primitives}

@@

The uniform $k$-CFA allocation strategy would make $eval$ in figure
\ref{fig:aam} a computable abstraction of reachable states, but not an
effective one. This is not the strategy that AAM, nor we,
recommend. We explain a succession of approximations to reach the
baseline analysis that we compare performance against. But first, we
need a fixpoint combinator to drive the succession of reduction
relations we define.

@@

We compute the semantics of a program by iterating the state
transition relation until a fixed point in the reachable states is
reached. We begin with Kleene's fixpoint combinator, ${\mathcal F}$,
presented in the frontier/accumulator style:

@@

A common technique to accelerate convergence in flow analyses is to share
a common store that is monotonically updated as interpretation
continues. In the ISWIM language with only unary operators, this makes
our 0-CFA quartic in the size of the program, which in one step drops the
complexity from intractable exponentiality to merely daunting
polynomiality.

@@

\begin{align*}
(cs, \msto) &\mathrel{\widehat{\machstep}} (cs', \msto') \\
\mbox{where } cs' &= \{c' \mid wn(c, \msto) \machstep wn(c, \msto^c), c \in cs\} \\
              \msto' &= \bigsqcup\{\msto^c \mid wn(c,\msto)\machstep wn(c', \msto^c), c \in cs\} \\
wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont}
\end{align*}
That is, the next set of states is what we get by stepping each
given state, without its store. The next store is the combination of all produced
stores. For simplicity, assume that stuck states step to themselves.

\subsection{Store-allocated results}
\label{sec:baselineeval}

@@

To achieve this, we instead allocate an address for the value position
when we create the continuation.  This address and the tail address
are both determined by the label of the application point, so the
space becomes linear and the overall complexity drops to cubic. Note
that this is a critical abstraction in languages with n-ary functions,
since otherwise the continuation space becomes ${\mathcal O}(n^n)$!
This amount of distinction between continuations is unnecessary
in practice.

@@

This extra store-allocation is effectively naming all intermediate
results, and thus the precision aligns with an analysis specialized to
ANF. We can also play a representation trick in 0-CFA and remove
$\menv$. In monovariant analyses, variables map to themselves, meaning
$\menv$ is effectively the identity function. That degeneration, in
turn, allows us to simply discard it from the analysis.

@@

\begin{align*}
(cs, \msto) &\mathrel{\widehat{\machstep}} (cs\cup cs', \msdiff(\msto)) \\
\mbox{ where } (cs', \msdiff) &= step^*(\varnothing, cs, \mtlst) \\
step^*(S, \varnothing, \msdiff) &= (S, \msdiff) \\
step^*(S, \{c\}\cup cs, \msdiff &= step^*(S\cup cs^*, cs, \msdiff^*) \\
cs^* &= \{c' \mid (c,\msto,\msdiff) \machstep (c',\msdiff^c) \} \\
\msdiff^* &= concat\{\msdiff^c \mid (c, \msto,\msdiff) \machstep (c', \msdiff^c)\}
\end{align*}
